Introduction to the \proglang{R} package \pkg{TDA}

### 2018-04-28 inst/doc/article.Rnw
version 1.6.2

%\VignetteIndexEntry{Introduction to TDA}
%% need no \usepackage{Sweave.sty}

@@

\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{verbatim,float}
\usepackage{graphicx,subfigure,url}
\usepackage{dsfont,bm,color,multirow}

%spacing for captions
\usepackage{sidecap}
\usepackage[font=small,skip=0pt]{caption}
\setlength{\captionmargin}{20pt}

@@

\author{Brittany T. Fasy, Jisu Kim, Fabrizio Lecci \and Cl\'ement Maria \\ \small{In collaboration with the CMU TopStat Group}}
\Plainauthor{Brittany Terese Fasy, Jisu Kim, Fabrizio Lecci, Cl\'ement Maria}

\title{Introduction to the \proglang{R} package \pkg{TDA}}
\Plaintitle{Introduction to the R package TDA}
\Abstract{
We present a short tutorial and introduction to using the \proglang{R} package 
\pkg{TDA}, which provides some tools for Topological Data Analysis.
In particular, it includes implementations of functions that, given some data, provide 
topological information about the underlying space, such as the distance function, the distance to a measure,
the kNN density estimator, the kernel density estimator, and the kernel distance.
The salient topological features of the sublevel sets (or superlevel sets) of these functions
can be quantified with persistent homology. We provide an \proglang{R} interface
for the efficient algorithms of the \proglang{C++} libraries \pkg{GUDHI}, \pkg{Dionysus} and \pkg{PHAT}, 
including a function for the persistent homology of the Rips filtration, 
and one for the persistent homology of sublevel
sets (or superlevel sets) of arbitrary functions evaluated over a grid of points.
The significance of the features in the resulting persistence diagrams can be 
analyzed with functions that implement the methods discussed in \cite{fasy2014statistical},
\cite{chazal2014stochastic} and \cite{chazal2014robust}.
The \proglang{R} package \pkg{TDA} also includes the implementation of an algorithm for density 
clustering, which  allows us to identify  the spatial organization of the probability mass associated to 
a density function and visualize it by means of a dendrogram, the cluster tree. 
}
\Keywords{Topological Data Analysis, Persistent Homology, Density Clustering}

\Address{
  Brittany T. Fasy\\
  Computer Science Department\\
  Tulane University\\
  Email: \email{brittany.fasy@alumni.duke.edu}\\

  Jisu Kim\\
  Department of Statistics\\
  Carnegie Mellon University\\
  Email: \email{jisuk1@andrew.cmu.edu}\\

  Cl\'ement Maria\\
  Geometrica Group\\
  INRIA Sophia Antipolis-M\'editerran\'ee\\
  Email: \email{clement.maria@inria.fr}\\

  Fabrizio Lecci\\
  Department of Statistics\\
  Carnegie Mellon University\\
  Email: \email{lecci@cmu.edu}\\

  TopStat Group\\
  Carnegie Mellon University\\
  Website: \url{http://www.stat.cmu.edu/topstat/} \\
  Email: \email{topstat@stat.cmu.edu}

@@

%There is a growing list of software available to compute persistent homology.
%\textsf{\textbf{javaPlex}}\footnote{https://code.google.com/p/javaplex/} is a
%Java software package developed by the Computational Topology workgroup at
%Stanford University;
%\textsf{\textbf{phom}}\footnote{
%http://cran.r-project.org/web/packages/phom/index.html} is an R package written
%by Andrew Tausz;
%\textsf{\textbf{PHAT}}\footnote{https://code.google.com/p/phat/} is a C++
%library developed at the Institute of Science and Technology Austria;
%\textsf{\textbf{Dionysus}}\footnote{http://www.mrzv.org/software/dionysus/} is
%a
%C++ library written and maintained by Dmitriy Morozov;
%finally, \textsf{\textbf{GUDHI}}\footnote{https://project.inria.fr/gudhi/} is
%new born project hosted by INRIA and whose goal is the development and
%implementation of new algorithms for geometric understanding in high
%dimensions.
%Preliminary results show that \textsf{\textbf{GUDHI}} outperforms its major
%competitors \textsf{\textbf{PHAT}} and \textsf{\textbf{Dionysus}}; see
%\cite{boissonnat2013compressed}.

This paper is devoted to the presentation of the \proglang{R} package \pkg{TDA},
which provides a user-friendly interface for the efficient algorithms of the 
\proglang{C++} libraries \pkg{GUDHI} \citep{gudhi_stpcoh}, \pkg{Dionysus} \citep{dionysus_C}, and \pkg{PHAT} \citep{phat_C}.

@@

<<eq2, echo=TRUE>>=
distance <- distFct(X = X, Grid = Grid)
@
\item Given a probability measure $P$, the distance to measure (DTM) is defined for each $y \in \mathbb{R}^d$ as
$$
d_{m_0}(y) =\sqrt{ \frac{1}{m_0}\int_0^{m_0} ( G_y^{-1}(u))^2 du},
$$
where
$G_y(t) = P( \Vert X-y \Vert \leq t)$ and $0<m_0<1$ is a smoothing parameter. 
The DTM can be seen as a smoothed version of the distance function.
For more details
see \cite{chazal2011geometric}.\\
Given $X=\{x_1, \dots, x_n\}$, the empirical version of the DTM is
$$
\hat d_{m_0}(y) = \sqrt{ \frac{1}{k} \sum_{x_i \in N_k(y)} \Vert x_i-y \Vert^2 },
$$
where $k= \lceil m_0 n \rceil$ and $N_k(y)$ is the 
set containing the $k$ nearest neighbors 
of $y$ among $x_1, \ldots, x_n$.
The DTM is computed for each point of the \code{Grid} with the following code:

@@

\cite{fasy2014statistical} and \cite{chazal2014robust} prove the validity of the bootstrap algorithm for kernel density estimators, distance to measure, and kernel distance, and use it in the framework of persistent homology.
The bootstrap algorithm is implemented in the function \code{bootstrapBand}, which provides the option of parallelizing the algorithm (\code{parallel=TRUE}) using the package \pkg{parallel}.
The following code computes a 90\% confidence band for $\mathbb{E}[\hat p_h]$, showed in Figure \ref{fig:eq10}.

@@

persp(Xseq, Yseq, matrix(band[["band"]][, 1], ncol = length(Yseq),
      nrow = length(Xseq)),
      zlim = c(0, max(band[["band"]][posYgrid, 2])),
      ylim = range(Yseq), xlab = "", ylab = "", zlab="",theta = 0,
      phi = 25, ltheta = 50, col = "pink", border = NA, d = 0.5,
      scale = FALSE, expand = 3, shade = 0.9, box = FALSE)

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Persistent Homology Over a Grid}
%%%%%%%%%%%%%%%%%%%%%%%
In this section, we describe how to use the \code{gridDiag} function to compute the persistent homology of sublevel (and superlevel) sets of the functions described in Section \ref{sec:distances}. 
The function \code{gridDiag} evaluates a given real valued function over a triangulated grid, constructs a filtration of simplices using the values of the function, and computes the persistent homology of the filtration. From version 1.2, \code{gridDiag} works in arbitrary dimension.
The core of the function is written in \proglang{C++} and the user can choose to compute persistence diagrams using either the \pkg{Dionysus} library or the \pkg{PHAT} library.

The following code computes the persistent homology of the superlevel sets \linebreak (\code{sublevel=FALSE}) of the kernel density estimator (\code{FUN=kde}, \code{h=0.3}) using the point cloud stored in the matrix~\code{X} from the previous example.
The same code would work for the other functions defined in Section \ref{sec:distances} (it is sufficient to replace \code{kde} and its smoothing parameter \code{h} with another function and the corresponding parameter). 
The function \code{gridDiag} returns an object of the class \code{"diagram"}.  
The other inputs are the features of the grid over which the \code{kde} is evaluated (\code{lim} and \code{by}), the smoothing parameter \code{h}, and a logical variable that indicates whether a progress bar should be printed (\code{printProgress}).

<<eq11, echo=TRUE, eval=TRUE>>=
Diag <- gridDiag(X = X, FUN = kde, h = 0.3, lim = cbind(Xlim, Ylim),
                 by = by, sublevel = FALSE, library = "Dionysus",
                 printProgress = FALSE)
@

We plot the data and the diagram, using the function \code{plot}, implemented as a standard \code{S3} method for objects of the class \code{"diagram"}. The following command produces the third plot in Figure \ref{fig:eq11b}.

<<eq11b, echo=TRUE, eval=FALSE>>=
plot(Diag[["diagram"]], band = 2 * band[["width"]],
     main = "KDE Diagram")
@

% PLOT STARTS
\begin{figure}[h!tb]
\setkeys{Gin}{width=5.5in, height=2in}
\begin{center}
<<eq11b, echo=FALSE, eval=TRUE, fig=TRUE, height=2, width=5.5>>=
par(mfrow = c(1, 3))
par(mai = c(0.3, 0.25, 0.3, 0.1))
plot(X, pch = 16, cex = 0.6, xlab = "", ylab = "", main = "Sample X")
persp(Xseq, Yseq,
      matrix(KDE, ncol = length(Yseq), nrow = length(Xseq)),
      xlab = "", ylab = "", zlab = "", theta = -20, phi = 35,
      ltheta = 50, col = 2, border = NA, main = "KDE", d = 0.5,
      scale = FALSE, expand = 3, shade = 0.9)
par(mai = c(0.55, 0.6, 0.3, 0.1))
plot(Diag[["diagram"]], band = 2 * band[["width"]],
     main = "KDE Diagram")
@
\end{center}
\caption{The plot on the right shows the persistence diagram of the superlevel sets of the KDE. Black points represent connected components and red triangles represent loops. 
The features are born at high levels of the density and die at lower levels. The pink 90\% confidence band separates significant features from noise.}
\label{fig:eq11b}
\end{figure}
% PLOT END

The option \code{band=2*band$width} produces a pink confidence band for the persistence diagram, using the confidence band constructed for the corresponding kernel density estimator in the previous section. 
The features above the band can be interpreted as representing significant homological features, while points in the band are not significantly different from noise.
The validity of the bootstrap confidence band for  persistence diagrams of KDE, DTM, and Kernel Distance derive from the \textit{Stability Theorem} \citep{chazal2012structure} and is discussed in detail in \cite{fasy2014statistical} and \cite{chazal2014robust}.

The function \code{plot} for the class \code{"diagram"} provide the options of rotating the diagram (\code{rotated=TRUE}), drawing the barcode in place of the diagram (\code{barcode=TRUE}), as well as other standard graphical options. See Figure \ref{fig:eq11c}.

% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=4.5in, height=2.25in}
\begin{center}
<<eq11c, echo=TRUE, eval=TRUE,  fig=TRUE, width=6, height=3>>=
par(mfrow = c(1, 2), mai = c(0.8, 0.8, 0.3, 0.1))
plot(Diag[["diagram"]], rotated = TRUE, band = band[["width"]],
     main = "Rotated Diagram")
plot(Diag[["diagram"]], barcode = TRUE,  main = "Barcode")
@
\end{center}
\caption{Rotated Persistence Diagram and Barcode}
\label{fig:eq11c}
\end{figure}
% PLOT END

@@

The \code{ripsDiag} function computes the persistence diagram of the Rips filtration built on top of a point cloud.
The user can choose to compute the Rips persistence diagram using either the \proglang{C++}  library \pkg{GUDHI}, or \pkg{Dionysus}.
\\The following code generates 60 points from two circles:

<<eq12, echo=TRUE, eval=TRUE>>=
Circle1 <- circleUnif(60)
Circle2 <- circleUnif(60, r = 2) + 3
Circles <- rbind(Circle1, Circle2)
@

@@

<<eq12b, echo=TRUE, eval=TRUE>>=
maxscale <- 5        # limit of the filtration
maxdimension <- 1    # components and loops
@

@@

<<eq12c, echo=TRUE, eval=TRUE>>=
Diag <- ripsDiag(X = Circles, maxdimension, maxscale,
                 library = "GUDHI", printProgress = FALSE)
@
Alternatively, using the option \code{dist="arbitrary"} in \code{ripsDiag()}, the input \code{X} can be an $n \times n$ matrix of distances.
This option is useful when the user wants to consider a Rips filtration constructed using an arbitrary distance and is currently only available for the option \code{library="Dionysus"}.\\

Finally we plot the data and the the diagram:\\
\phantom{o}
% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=4.6in, height=2.3in}
\begin{center}
<<eq12d, echo=TRUE, eval=TRUE,  fig=TRUE, height=2.8, width=5.6>>=
par(mfrow = c(1, 2), mai=c(0.8, 0.8, 0.3, 0.3))
plot(Circles, pch = 16, xlab = "",ylab = "")
plot(Diag[["diagram"]])
@
\end{center}
\caption{Rips persistence diagram. Black points represent connected components and red triangles represent loops.}
\label{fig:eq12d}
\end{figure}
% PLOT END

>>>>

%\VignetteIndexEntry{Introduction to TDA}
%% need no \usepackage{Sweave.sty

@@

\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{verbatim,float}
\usepackage{graphicx,subcaption,url}
\usepackage{bm,color}

%spacing for captions
\usepackage[font=small,skip=0pt]{caption}
\setlength{\captionmargin}{20pt}

@@

\author{Brittany T. Fasy, Jisu Kim, Fabrizio Lecci, Cl\'ement Maria, \and Vincent Rouvreau \\ \small{In collaboration with the CMU TopStat Group}}
\Plainauthor{Brittany Terese Fasy, Jisu Kim, Fabrizio Lecci, Cl\'ement Maria, Vincent Rouvreau}

\title{Introduction to the \proglang{R} package \pkg{TDA}}
\Plaintitle{Introduction to the R package TDA}
\Abstract{
We present a short tutorial and introduction to using the \proglang{R} package 
\pkg{TDA}, which provides some tools for Topological Data Analysis.
In particular, it includes implementations of functions that, given some data, provide 
topological information about the underlying space, such as the distance function, the distance to a measure,
the kNN density estimator, the kernel density estimator, and the kernel distance.
The salient topological features of the sublevel sets (or superlevel sets) of these functions
can be quantified with persistent homology. We provide an \proglang{R} interface
for the efficient algorithms of the \proglang{C++} libraries \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}, and \href{ https://bitbucket.org/phat-code/phat }{\pkg{PHAT}}, 
including a function for the persistent homology of the Rips filtration, 
and one for the persistent homology of sublevel
sets (or superlevel sets) of arbitrary functions evaluated over a grid of points.
The significance of the features in the resulting persistence diagrams can be 
analyzed with functions that implement the methods discussed in \cite{fasy2014statistical},
\cite{chazal2014stochastic} and \cite{chazal2014robust}.
The \proglang{R} package \pkg{TDA} also includes the implementation of an algorithm for density 
clustering, which allows us to identify the spatial organization of the probability mass associated to 
a density function and visualize it by means of a dendrogram, the cluster tree. 
}
\Keywords{Topological Data Analysis, Persistent Homology, Density Clustering}

\Address{
  Brittany T. Fasy\\
  Computer Science Department\\
  Montana State University\\
  Website: \url{ http://www.cs.montana.edu/brittany/ } \\
  Email: \email{brittany.fasy@alumni.duke.edu}\\

  Jisu Kim\\
  Department of Statistics\\
  Carnegie Mellon University\\
  Website: \url{ http://www.stat.cmu.edu/~jisuk/ } \\
  Email: \email{jisuk1@andrew.cmu.edu}\\

  Fabrizio Lecci\\
  Email: \email{fabrizio.lecci@gmail.com}\\

  Cl\'ement Maria\\
  School of Mathematics and Physics\\
  The University of Queensland\\
  Website: \url{ http://www-sop.inria.fr/members/Clement.Maria/ } \\
  Email: \email{clement.maria@inria.fr}\\

  Vincent Rouvreau\\
  GUDHI Group\\
  INRIA Saclay\\
  Email: \email{vincent.rouvreau@inria.fr}\\

  TopStat Group\\
  Carnegie Mellon University\\
  Website: \url{ http://www.stat.cmu.edu/topstat/ } \\
  Email: \email{topstat@stat.cmu.edu}

@@

%There is a growing list of software available to compute persistent homology.
%\textsf{\textbf{javaPlex}}\footnote{https://code.google.com/p/javaplex/} is a
%Java software package developed by the Computational Topology workgroup at
%Stanford University;
%\textsf{\textbf{phom}}\footnote{
%http://cran.r-project.org/web/packages/phom/index.html} is an R package written
%by Andrew Tausz;
%\textsf{\textbf{PHAT}}\footnote{https://bitbucket.org/phat-code/phat/} is a C++
%library developed at the Institute of Science and Technology Austria;
%\textsf{\textbf{Dionysus}}\footnote{http://www.mrzv.org/software/dionysus/} is
%a
%C++ library written and maintained by Dmitriy Morozov;
%finally, \textsf{\textbf{GUDHI}}\footnote{https://project.inria.fr/gudhi/} is
%new born project hosted by INRIA and whose goal is the development and
%implementation of new algorithms for geometric understanding in high
%dimensions.
%Preliminary results show that \textsf{\textbf{GUDHI}} outperforms its major
%competitors \textsf{\textbf{PHAT}} and \textsf{\textbf{Dionysus}}; see
%\cite{boissonnat2013compressed}.

This paper is devoted to the presentation of the \proglang{R} package \pkg{TDA},
which provides a user-friendly interface for the efficient algorithms of the 
\proglang{C++} libraries \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}} \citep{gudhi_stpcoh}, \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}} \citep{dionysus_C}, and \href{ https://bitbucket.org/phat-code/phat }{\pkg{PHAT}} \citep{phat_C}.

@@

<<eq2, echo=TRUE>>=
distance <- distFct(X = X, Grid = Grid)
@
\item Given a probability measure $P$, the distance to measure (DTM) is defined for each $y \in \mathbb{R}^d$ as
\[
d_{m0}(y) = \left(\frac{1}{m0}\int_0^{m0} ( G_y^{-1}(u))^{r} du\right)^{1/r},
\]
where
$G_y(t) = P( \Vert X-y \Vert \leq t)$, and $m0 \in (0,1)$ and $r \in [1,\infty)$ are tuning parameters. As \code{m0} increases, DTM function becomes smoother, so \code{m0} can be understood as a smoothing parameter. \code{r} affects less but also changes DTM function as well. The default value of \code{r} is \code{2}.
The DTM can be seen as a smoothed version of the distance function. See \citep[Definition 3.2]{chazal2011geometric} and \citep[Equation (2)]{ChazalMM2015} for a formal definition of the "distance to measure" function.

Given $X=\{x_1, \dots, x_n\}$, the empirical version of the DTM is
\[
\hat d_{m0}(y) = \left(\frac{1}{k} \sum_{x_i \in N_k(y)} \Vert x_i-y \Vert^{r}\right)^{1/r},
\]
where $k= \lceil m0 * n \rceil$ and $N_k(y)$ is the 
set containing the $k$ nearest neighbors 
of $y$ among $x_1, \ldots, x_n$.

For more details, see \citep{chazal2011geometric} and \citep{ChazalMM2015}.

The DTM is computed for each point of the \code{Grid} with the following code:

@@

\cite{fasy2014statistical} and \cite{chazal2014robust} prove the validity of the bootstrap algorithm for kernel density estimators, distance to measure, and kernel distance, and use it in the framework of persistent homology.
The bootstrap algorithm is implemented in the function \code{bootstrapBand}, which provides the option of parallelizing the algorithm (\code{parallel = TRUE}) using the package \pkg{parallel}.
The following code computes a 90\% confidence band for $\mathbb{E}[\hat p_h]$, showed in Figure \ref{fig:eq10}.

@@

persp(Xseq, Yseq, matrix(band[["band"]][, 1], ncol = length(Yseq),
      nrow = length(Xseq)),
      zlim = c(0, max(band[["band"]][posYgrid, 2])),
      ylim = range(Yseq), xlab = "", ylab = "", zlab = "", theta = 0,
      phi = 25, ltheta = 50, col = "pink", border = NA, d = 0.5,
      scale = FALSE, expand = 3, shade = 0.9, box = FALSE)

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Persistent Homology Over a Grid}
%%%%%%%%%%%%%%%%%%%%%%%
In this section, we describe how to use the \code{gridDiag} function to compute the persistent homology of sublevel (and superlevel) sets of the functions described in Section \ref{sec:distances}. 
The function \code{gridDiag} evaluates a given real valued function over a triangulated grid, constructs a filtration of simplices using the values of the function, and computes the persistent homology of the filtration. From version 1.2, \code{gridDiag} works in arbitrary dimension.
The core of the function is written in \proglang{C++} and the user can choose to compute persistence diagrams using either  the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat }{\pkg{PHAT}}.

The following code computes the persistent homology of the superlevel sets \linebreak (\code{sublevel = FALSE}) of the kernel density estimator (\code{FUN = kde}, \code{h = 0.3}) using the point cloud stored in the matrix~\code{X} from the previous example.
The same code would work for the other functions defined in Section \ref{sec:distances} (it is sufficient to replace \code{kde} and its smoothing parameter \code{h} with another function and the corresponding parameter). 
The function \code{gridDiag} returns an object of the class \code{"diagram"}.  
The other inputs are the features of the grid over which the \code{kde} is evaluated (\code{lim} and \code{by}), the smoothing parameter \code{h}, and a logical variable that indicates whether a progress bar should be printed (\code{printProgress}).

<<eqGrid1, echo=TRUE, eval=TRUE>>=
DiagGrid <- gridDiag(
    X = X, FUN = kde, h = 0.3, lim = cbind(Xlim, Ylim), by = by,
    sublevel = FALSE, library = "Dionysus", location = TRUE,
    printProgress = FALSE)
@

We plot the data and the diagram, using the function \code{plot}, implemented as a standard \code{S3} method for objects of the class \code{"diagram"}. The following command produces the third plot in Figure \ref{fig:eqGrid3}.

<<eqGrid2, echo=TRUE, eval=FALSE>>=
plot(DiagGrid[["diagram"]], band = 2 * band[["width"]],
     main = "KDE Diagram")
@

% PLOT STARTS
\begin{figure}[h!tb]
\setkeys{Gin}{width=5.5in, height=5.5in}
\begin{center}
<<eqGrid3, echo=FALSE, eval=TRUE, fig=TRUE, height=7, width=6>>=
par(mfrow = c(2, 2))
plot(X, pch = 16, cex = 0.6, xlab = "", ylab = "", main = "Sample X")
persp(Xseq, Yseq,
      matrix(KDE, ncol = length(Yseq), nrow = length(Xseq)),
      xlab = "", ylab = "", zlab = "", theta = -20, phi = 35,
      ltheta = 50, col = 2, border = NA, main = "KDE", d = 0.5,
      scale = FALSE, expand = 3, shade = 0.9)
plot(DiagGrid[["diagram"]], band = 2 * band[["width"]],
     main = "KDE Diagram")
one <- which(DiagGrid[["diagram"]][, 1] == 1)
plot(X, pch = 16, cex = 0.6, xlab = "", ylab = "", main = "Representative loop")
for (i in seq(along = one)) {
  points(DiagGrid[["birthLocation"]][one[i], , drop = FALSE], pch = 15, cex = 3,
      col = i)
  points(DiagGrid[["deathLocation"]][one[i], , drop = FALSE], pch = 17, cex = 3,
      col = i)
  for (j in seq_len(dim(DiagGrid[["cycleLocation"]][[one[i]]])[1])) {
    lines(DiagGrid[["cycleLocation"]][[one[i]]][j, , ], pch = 19, cex = 1, col = i)
  }
}

@
\end{center}
\caption{The plot on the right shows the persistence diagram of the superlevel sets of the KDE. Black points represent connected components and red triangles represent loops. 
The features are born at high levels of the density and die at lower levels. The pink 90\% confidence band separates significant features from noise.}
\label{fig:eqGrid3}
\end{figure}
% PLOT END

The option (\code{band = 2 * band[["width"]]}) produces a pink confidence band for the persistence diagram, using the confidence band constructed for the corresponding kernel density estimator in the previous section. 
The features above the band can be interpreted as representing significant homological features, while points in the band are not significantly different from noise.
The validity of the bootstrap confidence band for persistence diagrams of KDE, DTM, and Kernel Distance derive from the \textit{Stability Theorem} \citep{chazal2012structure} and is discussed in detail in \cite{fasy2014statistical} and \cite{chazal2014robust}.

The function \code{plot} for the class \code{"diagram"} provide the options of rotating the diagram (\code{rotated = TRUE}), drawing the barcode in place of the diagram (\code{barcode = TRUE}), as well as other standard graphical options. See Figure \ref{fig:eq11c}.

% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=4.5in, height=2.25in}
\begin{center}
<<eq11c, echo=TRUE, eval=TRUE, fig=TRUE, width=6, height=3>>=
par(mfrow = c(1, 2), mai = c(0.8, 0.8, 0.3, 0.1))
plot(DiagGrid[["diagram"]], rotated = TRUE, band = band[["width"]],
     main = "Rotated Diagram")
plot(DiagGrid[["diagram"]], barcode = TRUE, main = "Barcode")
@
\end{center}
\caption{Rotated Persistence Diagram and Barcode}
\label{fig:eq11c}
\end{figure}
% PLOT END

@@

The \code{ripsDiag} function computes the persistence diagram of the Rips filtration built on top of a point cloud.
The user can choose to compute the Rips filtration using either the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}} or \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}. Then for computing the persistence diagram from the Rips filtration, the user can use either the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}.
\\The following code generates 60 points from two circles:

<<eqRips1, echo=TRUE, eval=TRUE>>=
Circle1 <- circleUnif(60)
Circle2 <- circleUnif(60, r = 2) + 3
Circles <- rbind(Circle1, Circle2)
@

@@

<<eqRips2, echo=TRUE, eval=TRUE>>=
maxscale <- 5        # limit of the filtration
maxdimension <- 1    # components and loops
@

@@

<<eqRips3, echo=TRUE, eval=TRUE>>=
DiagRips <- ripsDiag(X = Circles, maxdimension, maxscale,
    library = c("GUDHI", "Dionysus"), location = TRUE, printProgress = FALSE)
@
Alternatively, using the option (\code{dist = "arbitrary"}) in \code{ripsDiag()}, the input \code{X} can be an $n \times n$ matrix of distances.
This option is useful when the user wants to consider a Rips filtration constructed using an arbitrary distance and is currently only available for the option (\code{library = "Dionysus"}).\\

Finally we plot the data and the diagram, as in Figure \ref{fig:eqRips4}.:\\
\phantom{o}
% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=6in, height=2in}
\begin{center}
<<eqRips4, echo=FALSE, eval=TRUE, fig=TRUE, height=3, width=9>>=
par(mfrow = c(1, 3), mai=c(0.8, 0.8, 0.3, 0.3))
plot(Circles, pch = 16, xlab = "",ylab = "", main = "Two Circles")
plot(DiagRips[["diagram"]], main = "Rips persistence diagram")
one <- which(DiagRips[["diagram"]][, 1] == 1 &
    DiagRips[["diagram"]][, 3] - DiagRips[["diagram"]][, 2] > 0.5)
plot(Circles, col = 2, main = "Representative loop")
for (i in seq(along = one)) {
  for (j in seq_len(dim(DiagRips[["cycleLocation"]][[one[i]]])[1])) {
    lines(DiagRips[["cycleLocation"]][[one[i]]][j, , ], pch = 19, cex = 1, col = i)
  }
}
@
\end{center}
\caption{Rips persistence diagram. Black points represent connected components and red triangles represent loops.}
\label{fig:eqRips4}
\end{figure}
% PLOT END

%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Alpha Complex Persistence Diagram}
%%%%%%%%%%%%%%%%%%%%%%%%%%
For a finite set of points $X \subset \mathbb{R}^{d}$, the {\em Alpha complex} $Alpha(X,s)$ is a simplicial subcomplex of the Delaunay complex of $X$ consisting of simplices of circumradius less than or equal to $\sqrt{s}$. For each $u \in X$, let $V_{u}$ be its Voronoi cell, i.e. $V_{u}=\{x\in \mathbb{R}^{d}:\,d(x,u)\leq d(x,v)\text{ for all }v\in X\}$, and $B_{u}(r)$ be the closed ball with center $u$ and radius $r$. Let $R_{u}(r)$ consists of be the intersection of earh ball of radius $r$ with the voronoi cell of $u$, i.e. $R_{u}(r)=B_{u}(r)\cap V_{u}$. Then $Alpha(X,s)$ is defined as
\[
Alpha(X,r)=\left\{ \sigma\subset X:\,\underset{u\in\sigma}{\bigcap}R_{u}(\sqrt{s})\neq\emptyset\right\} .
\]
See \citep[Section 3.4]{edelsbrunner2010computational} and \citep{gudhi:AlphaComplex}. The sequence of Alpha complexes obtained by gradually increasing the parameter $s$ creates an Alpha complex filtration.

The \code{alphaComplexDiag} function computes the Alpha complex filtration built on top of a point cloud, using the \proglang{C++} library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}. Then for computing the persistence diagram from the Alpha complex filtration, the user can use either the C++ library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}.

We first generate 30 points from a circle:

<<eqAlphaComplex1, echo=TRUE, eval=TRUE>>=
X <- circleUnif(n = 30)
@

and the following code compute the persistence diagram of the alpha complex filtration using the point cloud \code{X}, with printing its progress (\code{printProgress = FALSE}). The function \code{alphaComplexDiag} returns an object of the class \code{"diagram"}.

<<eqAlphaComplex2, echo=TRUE, eval=TRUE>>=
# persistence diagram of alpha complex
DiagAlphaCmplx <- alphaComplexDiag(
    X = X, library = c("GUDHI", "Dionysus"), location = TRUE,
    printProgress = TRUE)
@

And we plot the diagram in Figure \ref{fig:eqAlphaComplex3}.

% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=4in, height=2.3in}
\begin{center}
<<eqAlphaComplex3, echo=TRUE, eval=TRUE, fig=TRUE, height=3, width=8>>=
# plot
par(mfrow = c(1, 2))
plot(DiagAlphaCmplx[["diagram"]], main = "Alpha complex persistence diagram")
one <- which(DiagAlphaCmplx[["diagram"]][, 1] == 1)
one <- one[which.max(
    DiagAlphaCmplx[["diagram"]][one, 3] - DiagAlphaCmplx[["diagram"]][one, 2])]
plot(X, col = 1, main = "Representative loop")
for (i in seq(along = one)) {
  for (j in seq_len(dim(DiagAlphaCmplx[["cycleLocation"]][[one[i]]])[1])) {
    lines(DiagAlphaCmplx[["cycleLocation"]][[one[i]]][j, , ], pch = 19,
	    cex = 1, col = i + 1)
  }
}
par(mfrow = c(1, 1))
@
\end{center}
\caption{Persistence diagram of Alpha complex. Black points represent connected components and red triangles represent loops.}
\label{fig:eqAlphaComplex3}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Persistence Diagram of Alpha Shape}
%%%%%%%%%%%%%%%%%%%%%%%%%%
The {\em Alpha shape complex} $S(X,\alpha)$ is the polytope with its boundary consisting of $\alpha$-exposed simplices, where a simplex $\sigma$ is $\alpha$-exposed if there is an open ball $b$ of radius $\alpha$ such that $b \cap X = \emptyset$ and $\partial b \cap X = \sigma$. Suppose $\mathbb{R}^{d}$ is filled with ice cream, then consider scooping out the ice cream with sphere-shaped spoon of radius $\alpha$ without touching the points $X$. $S(X,\alpha)$ is the remaining polytope with straightening round surfaces. See \citep{fischer2005} and \citep{edelsbrunner1994}. The sequence of Alpha shape complexes obtained by gradually increasing the parameter $\alpha$ creates an Alpha shape complex filtration.

The \code{alphaShapeDiag} function computes the persistence diagram of the Alpha shape filtration built on top of a point cloud in 3 dimension, using the C++ library \href{ https://project.inria.fr/gudhi/software }{\pkg{GUDHI}}. Then for computing the persistence diagram from the Alpha shape filtration, the user can use either the C++ library \href{ https://project.inria.fr/gudhi/software/ }{\pkg{GUDHI}}, \href{ http://www.mrzv.org/software/dionysus/ }{\pkg{Dionysus}}, or \href{ https://bitbucket.org/phat-code/phat/ }{\pkg{PHAT}}. Currently the point data cloud should lie in 3 dimension.

We first generate 30 points from a cylinder:

<<eqAlphaShape1, echo=TRUE, eval=TRUE>>=
n <- 30
X <- cbind(circleUnif(n = n), runif(n = n, min = -0.1, max = 0.1))
@

and the following code compute the persistence diagram of the alpha shape filtration using the point cloud \code{X}, with printing its progress (\code{printProgress = TRUE}). The function \code{alphaShapeDiag} returns an object of the class \code{"diagram"}.

<<eqAlphaShape2, echo=TRUE, eval=TRUE>>=
DiagAlphaShape <- alphaShapeDiag(
    X = X, maxdimension = 1, library = c("GUDHI", "Dionysus"), location = TRUE,
    printProgress = TRUE)
@

And we plot the diagram and first two dimension of data in Figure \ref{fig:eqAlphaShape3}.

% PLOT START
\begin{figure}[h!tb]
\setkeys{Gin}{width=4in, height=2.3in}
\begin{center}
<<eqAlphaShape3, echo=TRUE, eval=TRUE, fig=TRUE, height=3, width=8>>=
par(mfrow = c(1, 2))
plot(DiagAlphaShape[["diagram"]])
plot(X[, 1:2], col = 2, main = "Representative loop of alpha shape filtration")
one <- which(DiagAlphaShape[["diagram"]][, 1] == 1)
one <- one[which.max(
    DiagAlphaShape[["diagram"]][one, 3] - DiagAlphaShape[["diagram"]][one, 2])]
for (i in seq(along = one)) {
  for (j in seq_len(dim(DiagAlphaShape[["cycleLocation"]][[one[i]]])[1])) {
    lines(
        DiagAlphaShape[["cycleLocation"]][[one[i]]][j, , 1:2], pch = 19,
        cex = 1, col = i)
  }
}
par(mfrow = c(1, 1))
@
\end{center}
\caption{Persistence diagram of Alpha shape. Black points represent connected components and red triangles represent loops.}
\label{fig:eqAlphaShape3}
\end{figure}
